---
title: "Final Project"
author: "Joshua Ramirez-Capistran"
format: html
editor: visual
---

# Data

```{r}
library(tidyverse)

border_crossings <- read_csv("data/Border_Crossing_Entry_Data.csv")
```

Background: I will be using the Border Crossing Entry Data, that I found using data.gov, which is a credible source for data. The variables in the data set would be Port Name, State, Port Code, Border, Date, Measure, Value, Latitude, Longitude, Point.

Numerical Continuous: Latitude, Longitude, Point

Numerical Discrete: Date, Value

Categorical Ordinal: Port Code

Categorical Nominal: Port Name, State, Border, Measure

These variables will show all sorts of things, like locations, entry point, vehicle type (measure), and the amount of people in each vehicle (value).

# Questions

1)  What is the distribution of the number of crossings at U.S. border entry points?

2)  How does the number of crossings (value) differ between the U.S.-Canada and U.S.-Mexico borders?

3)  What is the geographic distribution of U.S. border entry points based on their latitude and longitude?

4)  How do geographic location (latitude and longitude) and border type (U.S.-Canada vs. U.S.-Mexico) predict the number of crossings at U.S. border entry points?

# Descriptive Analysis

Histogram:

```{r}
border_crossings |>
  ggplot(aes(x = Value)) +
  geom_histogram()
```

Boxplot:

```{r}
border_crossings |>
  ggplot(aes(x = Border)) +
  geom_boxplot()

```

Scatterplot:

```{r}
border_crossings |>
  ggplot(aes(y = Longitude, x = Latitude)) +
  geom_point()

```

Calculate mean, standard deviation, minimum and maximum of at least one numeric variable by a group

```{r}
border_crossings |>
  group_by(Border) |>
  summarize(mean = mean(Value, na.rm = TRUE),
            median = median(Value, na.rm = TRUE),
            max = max(Value, na.rm = TRUE),
            min = min(Value, na.rm = TRUE),
            sd = sd(Value, na.rm = TRUE))
```

# Inferential Statistics

Linear Regression:

```{r}
model <- lm(Value ~ Latitude + Longitude + Border, 
            data = border_crossings)

summary(model)

```

Analysis:

R-squared = 0.054

The model explains \~5.4% of the variation in value (not very strong — but common for geographic/social data)

All variables are statistically significant (p \< 0.05)

Border type and geographic location (latitude & longitude) significantly predict the number of crossings.

There are more crossings at the US-Mexico border.

Locations further north and east tend to have fewer crossings

# Conclusion

For my analysis, I explored patterns in the U.S. Border Crossing Entry Data by examining the distribution and relationships of border traffic volume (value). I first visualized the distribution of crossing counts with a histogram, which showed that most entry points experience a low to moderate number of crossings and few high traffic locations. A boxplot was used to compare crossings by border which revealed that the U.S.-Mexico border essentially has higher crossing counts than the U.S. - Canada Border. A scatterplot was used for the entry point locations (latitude vs. longitude) and showed the geographic spread of ports, with clusters representing each countries border. I also summarized the number of crossings by border type, calculating mean, standard deviation, minimum and maximum values. It confirmed that U.S. - Mexico border crossings were much more than the U.S. - Canada Border.

After, I did a linear regression model using latitude, longitude, and border type. The regression results showed that the crossings are significantly higher at the U.S. - Mexico border compared to the U.S. - Canada border. Although the model explains about 5.4% of the variation in crossing counts, all variables were statistically significant, stating that geographic location and border type do play roles and influence crossings at U.S. entry points.
